The brightness and spatial distributions of terrestrial radio sources

### 2013-04-02 articles/Article-RFI-distribution/article-rfi-distribution.tex
Work on referee comments

We need the complement of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance. Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2$, with $N$ the total number of sources. Because $r$ is restricted, $N$ is finite.

@@

A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. The original model is based on urban areas, but corrections are given by Hata for sub-urban areas with a lower population density and for open areas. Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss over land:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in meters; $h_m$ the height of the receiving antenna in meters; $r$ the distance between the antennas in meters; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation} \label{eq:zeta-definition}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{Ig}{4\pi \zeta r^\eta},
\end{equation}
with $\eta = 4.49 - 0.655 \log_{10} h_b$, one finds the rate density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(rho*%pi*r^2,S),solve(S(r)=S,r);
\begin{align} \label{eq:frequency-density-with-propagation}
f_p(S)
= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{Ig }{4\pi\zeta S} \right)^{2/\eta} \right]
= \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{4\pi\zeta S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram. According to Hata's model, this is valid for the range 30--200 m. The trend of the slope will not continue indefinitely when increasing the height further. Instead it will converge to a $-2$ slope, which corresponds to free-space propagation.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

@@

\begin{figure}
\begin{center}\hspace{-5mm}\includegraphics[width=8cm]{img/plot-rayleigh-and-rfi-trimmed}
\caption{The Rayleigh and power-law distributions in a log-log plot. The power-law distribution (Eq.~\eqref{eq:two-dimensional-distribution}) has a constant slope of -2. The slope of the Rayleigh distribution in the limit of the origin is 1. Its maximum occurs where the amplitude value equals its mode $\sigma$, which is 1 in this example. For higher amplitudes, its slope decreases exponentially.}
\label{fig:rayleigh-and-rfi-distributions}
\end{center}
\end{figure}

@@

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial. The histogram can be numerically estimated by drawing complex samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by scaling and inverting the cumulative rate density function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, one needs to assume that there are no samples beyond some limiting distance $r_L$ to invert the cumulative function. With this assumption, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
\begin{equation} \label{eq:drawing-rfi-distribution}
S_{\textrm{RFI}} \leftarrow \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
\end{equation}
where $x_u$ and $y_u$ are independently drawn from a uniform distribution between 0 and 1. The result is an amplitude sample $S_{\textrm{RFI}}$, with $S_{\textrm{RFI}} \ge Ig/r_L^2$, that follows a power-law distribution. The corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
\begin{equation} \label{eq:drawing-combined-distribution}
S \leftarrow v_n + w_n i + \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
\end{equation}
with $v_n, w_n \sim N(\mu=0; \sigma)$. The final amplitude sample $S_A$ can be calculated with 
\begin{equation} \label{eq:drawing-amplitudes}
S_A \leftarrow |S| = \sqrt{\textrm{real}^2({S})+\textrm{imag}^2({S})}.
\end{equation}
An example of distribution curves for various settings of $Ig/r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.

If we assume that $I$, the average source strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-combined-distribution} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram. Although calibration could in theory solve $g$, almost all sources will come in through the far-sidelobes of the phased-array beam, and finding the expected values for the gains is therefore hard. The effects of these parameters on the histogram will be further discussed in \S\ref{sec:parameter-variability}.

Fig.~\ref{fig:rayleigh-and-rfi-combined} shows that the left side of the graph corresponds with the original Rayleigh distribution. From the right side of the histogram, we are able to estimate $Ig/r_L^2$, as the placing of the power-law curve is independent of the Rayleigh distribution for large $S$. Methods to constrain the parameters of the RFI distribution will be discussed in \S\ref{sec:rfi-distribution-constraints}. The curves in Fig.~\ref{fig:rayleigh-and-rfi-combined} are from histograms of samples that are all contaminated by both noise and RFI. In practice, we can not make this distinction accurately, as RFI detectors have a limited accuracy. Consequently, the observed histogram will be the sum of two types of histograms: the first histogram being contaminated by both noise and RFI, the second one only by noise.

For completeness we show the formula for sampling apparent source strengths that include propagation effects according to Hata's model, which can be used for sampling realistic source amplitudes in simulations. This can be derived by integration, scaling and inversion of the rate density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
\begin{equation} \label{eq:sample-with-propagation}
\tilde{S}_{\textrm{RFI}} \leftarrow \frac{I g}{\zeta x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
\end{equation}
Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 \le x_u,y_u < 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.

\subsection{Parameter variability} \label{sec:parameter-variability}
In reality, the parameters $\rho$, $I$ and $g$, which are the source density per unit area, source strength and instrumental gain respectively, will not be constant, but can change over time and frequency. Therefore, they act like stochastic variables. However, since each specific value for these parameters produces a power law, the combined distribution will still show a power law, as long as the parameters follow a distribution that is steep at high amplitudes (in $\log$--$\log$ space), such as a Gaussian or uniform distribution.

One instrumental effect that is absorbed in $g$ is the frequency response of the instrument, i.e., the antenna response in combination with the band-pass of the analogue and digital filters. Because the data that are analysed in Sect.~\ref{sec:dist-results} have initially not been band-pass calibrated, the instrumental response is not uniform over frequency. We determined that the variation due to the band-pass is about one order of magnitude for the low-band antennas (LBA, 30.1--77.5 MHz) and about a factor of two for the high-band antennas (HBA, 115.0--163.3 MHz).

The effect of the band-pass on the data distribution is consequently limited to one order of magnitude or less. If the RFI sources have an apparent power-law distribution with a certain exponent, the distribution will have a feature at low amplitudes due to the frequency response, but a power law with equal exponent will still appear when the distribution is observed over a wide enough range.

@@

One should note that fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}. However, because of our enormous sample size, which allows fitting the line over a large interval, the estimator will be sufficiently accurate for our purpose. Nevertheless, we will additionally calculate a maximum likelihood estimator for comparison. The maximum-likelihood estimator for the exponent in a power-law distribution is given by the Hill estimator $\hat \alpha_H$ \citep{hill-estimator, power-law-distribution}, defined as:
\begin{equation} \label{eq:hill}
 \hat \alpha_H = 1 + N \left(\sum\limits_{i=1}^{N} \ln \frac{x_i}{x_\textrm{min}} \right)^{-1},
\end{equation}
with $N$ the number of samples and $x_i$ for $0 < i \le N$ the samples that follow a power law with lower bound $x_\textrm{min}$.
However, this estimator assumes the distribution is not cut-off at a high point. In our case, the distribution is cut-off at both sides, for example because of the limit of the analogue-digital converter (see \S\ref{sec:rfi-distribution-constraints}). Therefore, using this estimator will result in an estimate that is steeper (i.e., more negative) than the actual distribution. Moreover, the estimator requires reading and processing all data points, and because of the size of the data sets this is somewhat impractical. However, we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ is the bin rate. Because our bin size is small and the distribution is well sampled, the resulting estimate will be close to the normal Hill estimator. This avoids the need to iterate over all data.

@@

$S_1$ and $S_2$ are selected by visually inspection of the histogram. Assume that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}, and that the effect of the Rayleigh component on the histogram $> S_1$ is negligible. The hypothetical upper limit $S_U$ of the distribution can now be found, i.e., the highest amplitude that would be observed when the distribution follows the power law up to that point, by solving
\begin{equation} \label{eq:upper-limit-constrained}
 \int\limits_{S_1}^{S_U} h(S) dS = N_\textrm{part}.
\end{equation}
In practice, the observed histogram will break down beyond some amplitude because of several reasons. First and most importantly, the samples themselves are digitized with an analogue-to-digital converter (ADC) with limited range. Second, we observe for a limited time and the rate count is discrete. Because the chance of finding a sample with a very high intensity is low, it is unlikely to observe samples beyond some amplitude within the finite observing time. Finally, under the assumption of a uniform spatial distribution of RFI transmitters, samples with very high amplitude would have to be produced by transmitters that are very close to the telescope. However, it is likely that the uniform spatial distribution of transmitters will break down at closer distances.

Solving Eq.~\eqref{eq:upper-limit-constrained} results in
\begin{equation} \label{eq:upper-limit}
S_U = \sqrt[\alpha+1]{\frac{\alpha+1}{\beta} N_\textrm{part} + S_1^{\alpha+1}}.
\end{equation}
In some cases there will be no solution for $S_U$. This situation arises because the integral of the distribution function is finite even when $S_U \rightarrow \infty$. If $N_\textrm{part}$ exceeds the integrated value of the fitted distribution function, there is no solution. This can happen when the empirical distribution is not limited at the high end, or when it contains features that are not in the model.

@@

With estimates of $\alpha$, $\beta$, $S_L$ and $S_U$, one has obtained a parametrization of the RFI distribution. As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is $Ig/r^2$ -- assuming free space propagation for the moment -- and therefore $S_L = Ig/r^2$. For Hata's propagation model, the more generic solution $S_L = Ig / (\zeta r^\alpha)$ is found. This value represents the apparent brightness of the sources that are the furthest away from the telescope. With a fully parametrized distribution of the effect of RFI sources, an empirical model for RFI effects can be made. Moreover, one can calculate $E(S_R)$, the expected apparent strength of RFI:
\begin{align} \label{eq:expected-value-rfi-def}
E(S_R) & = \frac{1}{N_{LU}}\int\limits_{S_L}^{S_U} \beta S^\alpha S dS
 = \frac{\beta}{N_{LU}} \left[ \frac{1}{\alpha+2} S^{\alpha+2} \right]_{S_L}^{S_U} \\
 & = \frac{\beta \left( S_U^{\alpha+2} - S_L^{\alpha+2} \right) }{N_{LU} \left( \alpha+2\right) }
\end{align}
Here, $N_{LU}$ is the number of samples between $S_L$ and $S_U$ after normalizing for the bin size:
\begin{align} \label{eq:n-between-sl-su}
N_{LU} = \int\limits_{S_L}^{S_U} \beta S^\alpha dS.
\end{align}
Substitution and simplification of these two results in
\begin{align} \label{eq:expected-value-rfi}
 E(S_R) & = \frac
{\left( S_U^{\alpha+2} - S_L^{\alpha+2}\right)\left(\alpha+1\right)} 
{\left(S_U^{\alpha+1} - S_L^{\alpha+1}\right) \left(\alpha+2 \right)}.
\end{align}
This is essentially the average flux density that is caused by RFI without using RFI detection or excision algorithms. $E(S_R)$ has the same units as $S_L$ and $S_U$, thus after calibration (see \S\ref{sec:calibration}) could be given in Jy. In practice, the increase of data noise after correlation is much less severe because of RFI flagging, which excises a part of the RFI. One can assume that all RFI above a certain power level is found by the detector. Since modern RFI detection algorithms can find all RFI that is detectable ``by eye'' \citep{post-correlation-rfi-classification}, this power level will be near the level of the noise mode. In
\citet{lofar-radio-environment}
the false-positives rate for the AOFlagger is estimated to be 0.5\%. An estimate of $S_d$, the average lower limit of detected RFI, can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the real number (true positives) of RFI samples. Therefore, the limit is calculated similar to Eq.~\eqref{eq:lower-limit-1}, where the term $N_\textrm{part} - N_\textrm{total}$ needs to be replaced with $N_\textrm{RFI}$, which equals the total number of samples detected as RFI minus the 0.5\% false positives.

@@

To detect RFI, the AOFlagger \citep{LOFAR-RFI-pipeline, scale-invariant-rank-operator} was used. However, since this flagger is partly amplitude-based, it is likely that low-level RFI will leak through the detector. Since it is also low-level RFI we are interested in, we will analyse unflagged data and the RFI classified data.

>>>>

We need $F_{\textrm{distance} \ge r}(r)$, the complement of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance. Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2$, with $N$ the total number of sources. Because $r$ is restricted, $N$ is finite.

@@

A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. The original model is based on urban areas, but corrections are given by Hata for suburban areas with a lower population density and for open areas. Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss between two ground-based antennas:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in meters; $h_m$ the height of the receiving antenna in meters; $r$ the distance between the antennas in meters; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \frac{1}{10} 10^{L_p} = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation} \label{eq:zeta-definition}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{Ig}{4\pi \zeta r^\eta},
\end{equation}
with $\eta = 4.49 - 0.655 \log_{10} h_b$ as seen in Eq.~\eqref{eq:propagation-loss-factor}, one finds the rate density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(rho*%pi*r^2,S),solve(S(r)=S,r);
\begin{align} \label{eq:frequency-density-with-propagation}
f_p(S)
= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{Ig }{4\pi\zeta S} \right)^{2/\eta} \right]
= \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{4\pi\zeta S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram. According to Hata's model, this is valid for the range 30--200 m. The trend of the slope will not continue indefinitely when increasing the height further. Instead it will converge to a $-2$ slope, which corresponds to free-space propagation.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

@@

\begin{figure}
\begin{center}\hspace{-5mm}\includegraphics[width=8cm]{img/plot-rayleigh-and-rfi-trimmed}
\caption{The Rayleigh and power-law distributions in a log-log plot. The power-law distribution (Eq.~\eqref{eq:two-dimensional-distribution}) has a constant slope of -2. Its integral is not equal to 1 in this example, because some form of flattening would be required to make this possible. The slope of the Rayleigh distribution in the limit of the origin is 1. Its maximum occurs where the amplitude value equals its mode $\sigma$, which is 1 in this example. For higher amplitudes, its slope decreases exponentially.}
\label{fig:rayleigh-and-rfi-distributions}
\end{center}
\end{figure}

@@

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial. The histogram can be numerically estimated by drawing complex samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by scaling and inverting the cumulative rate density function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, one needs to assume that there are no samples beyond some limiting distance $r_L$ to invert the cumulative function. With this assumption, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
\begin{equation} \label{eq:drawing-rfi-distribution}
S_{\textrm{RFI}} \leftarrow \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
\end{equation}
where $0 < x_u, y_u \le 1$ are independently drawn from a uniform distribution. The result is an amplitude sample $S_{\textrm{RFI}}$, with $S_{\textrm{RFI}} \ge Ig/r_L^2$, that follows a power-law distribution. The corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
\begin{equation} \label{eq:drawing-combined-distribution}
S \leftarrow v_n + w_n i + \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
\end{equation}
with $v_n, w_n \sim N(\mu=0; \sigma)$. The final amplitude sample $S_A$ can be calculated with 
\begin{equation} \label{eq:drawing-amplitudes}
S_A \leftarrow |S| = \sqrt{\textrm{real}^2({S})+\textrm{imag}^2({S})}.
\end{equation}
An example of distribution curves for various settings of $Ig/r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.

If we assume that $I$, the average source strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-combined-distribution} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram. Although calibration could in theory solve $g$, almost all RFI sources will come in through the far-sidelobes of the beam, and finding the expected values for the gains is therefore hard. The effects of these parameters on the histogram will be further discussed in \S\ref{sec:parameter-variability}.

%Fig.~\ref{fig:rayleigh-and-rfi-combined} shows that the left side of the graph corresponds with the original Rayleigh distribution. From the right side of the histogram, we are able to estimate $Ig/r_L^2$, as the placing of the power-law curve is independent of the Rayleigh distribution for large $S$. Methods to constrain the parameters of the RFI distribution will be discussed in \S\ref{sec:rfi-distribution-constraints}. The curves in Fig.~\ref{fig:rayleigh-and-rfi-combined} are from histograms of samples that are all contaminated by both noise and RFI. In practice, we can not make this distinction accurately, as RFI detectors have a limited accuracy. Consequently, the observed histogram will be the sum of two types of histograms: the first histogram being contaminated by both noise and RFI, the second one only by noise.

For completeness we show the formula for sampling apparent source strengths that include propagation effects according to Hata's model, which can be used for sampling realistic source amplitudes in simulations. This can be derived by integration, scaling and inversion of the rate density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
\begin{equation} \label{eq:sample-with-propagation}
\tilde{S}_{\textrm{RFI}} \leftarrow \frac{I g}{\zeta x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
\end{equation}
Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 < x_u,y_u \le 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.

\subsection{Parameter variability} \label{sec:parameter-variability}
In reality, the parameters $\rho$, $I$ and $g$, which are the source density per unit area, source strength and instrumental gain respectively, will not be constant, but can change over time and frequency. Therefore, they are stochastic variables. However, since each specific value for these parameters produces a power law, the combined distribution will still show a power law, as long as the parameters follow a distribution that is steep at high amplitudes (in $\log$--$\log$ space), such as a Gaussian or uniform distribution.

One instrumental effect that is absorbed in $g$ is the frequency response of the instrument, i.e., the antenna response in combination with the band-pass of the analogue and digital filters. Because the data that are analysed in Sect.~\ref{sec:dist-results} have initially not been band-pass calibrated, the instrumental response is not uniform over frequency. We determined that the gain variation due to the band-pass is about one order of magnitude for the low-band antennas (LBA, 30.1--77.5 MHz) and about a factor of two for the high-band antennas (HBA, 115.0--163.3 MHz). The frequency dependency of the gains due to the band-pass will consequently smooth the data in the brightness histogram in horizontal direction by one order of magnitude or less.

@@

Fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}. However, because of our enormous sample size, which allows fitting the line over a large interval, the estimator will be sufficiently accurate for our purpose. Nevertheless, we will additionally calculate a maximum-likelihood estimator for comparison. The maximum-likelihood estimator for the exponent in a power-law distribution is given by the Hill estimator $\hat \alpha_H$ \citep{hill-estimator, power-law-distribution}, defined as:
\begin{equation} \label{eq:hill}
 \hat \alpha_H = 1 + N \left(\sum\limits_{i=1}^{N} \ln \frac{x_i}{x_\textrm{min}} \right)^{-1},
\end{equation}
with $N$ the number of samples and $x_i$ for $0 < i \le N$ the samples that follow a power law with lower bound $x_\textrm{min}$.
However, this estimator assumes the distribution is not cut-off at a high point. In our case, the distribution is cut-off at both sides, for example because of the limit of the analogue-digital converter (see \S\ref{sec:rfi-distribution-constraints}). Therefore, using this estimator will result in an estimate that is steeper (i.e., more negative) than the actual distribution.

@@

$S_1$ and $S_2$ are selected by visual inspection of the histogram. Assume that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}, and that the effect of the Rayleigh component on the histogram $> S_1$ is negligible. The hypothetical upper limit $S_U$ of the distribution can now be found, i.e., the highest amplitude that would be observed when the distribution follows the power law up to that point, by solving
\begin{equation} \label{eq:upper-limit-constrained}
 \int\limits_{S_1}^{S_U} h(S) dS = N_\textrm{part}.
\end{equation}
In practice, the observed histogram will break down beyond some amplitude because of several reasons. First and most importantly, the samples themselves are digitized with an analogue-to-digital converter (ADC) with limited range. Second, we observe for a limited time and the rate count is discrete. Because the chance of finding a sample with a very high intensity is low, it is unlikely to observe samples beyond some amplitude within the finite observing time. Finally, under the assumption of a uniform spatial distribution of RFI transmitters, samples with very high amplitude would have to be produced by transmitters that are very close to the telescope. However, it is likely that the uniform spatial distribution of transmitters will break down at closer distances.

Solving Eq.~\eqref{eq:upper-limit-constrained} results in
\begin{equation} \label{eq:upper-limit}
S_U = \sqrt[\alpha+1]{\frac{\alpha+1}{\beta} N_\textrm{part} + S_1^{\alpha+1}}.
\end{equation}
In some cases there will be no solution for $S_U$. This can happen when the empirical distribution is not limited at the high end, or when it contains features that are not in the model.

@@

With estimates of $\alpha$, $\beta$, $S_L$ and $S_U$, one has obtained a parametrization of the RFI distribution. As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is $Ig/(4\pi r^2)$ -- assuming free space propagation for the moment -- and therefore $S_L = Ig/(4\pi r^2)$. For Hata's propagation model, the more generic solution $S_L = Ig / (4\pi \zeta r^\alpha)$ is found. This value represents the apparent brightness of the sources that are the furthest away from the telescope. With a fully parametrized distribution of the effect of RFI sources, an empirical model for RFI effects can be made. Moreover, one can calculate $E(S_R)$, the expected apparent strength of RFI:
\begin{align} \label{eq:expected-value-rfi-def}
E(S_R) & = \frac{1}{N_{LU}}\int\limits_{S_L}^{S_U} \beta S^\alpha S dS
 = \frac{\beta}{N_{LU}} \left[ \frac{1}{\alpha+2} S^{\alpha+2} \right]_{S_L}^{S_U} \\
 & = \frac{\beta \left( S_U^{\alpha+2} - S_L^{\alpha+2} \right) }{N_{LU} \left( \alpha+2\right) }
\end{align}
Here, $N_{LU}$ is the number of samples between $S_L$ and $S_U$ after normalizing for the bin size:
\begin{align} \label{eq:n-between-sl-su}
N_{LU} = \int\limits_{S_L}^{S_U} \beta S^\alpha dS.
\end{align}
Substitution and simplification of these two results in
\begin{align} \label{eq:expected-value-rfi}
 E(S_R) & = \frac
{\left( S_U^{\alpha+2} - S_L^{\alpha+2}\right)\left(\alpha+1\right)} 
{\left(S_U^{\alpha+1} - S_L^{\alpha+1}\right) \left(\alpha+2 \right)}.
\end{align}
This is essentially the average flux density that is caused by RFI without using RFI detection or excision algorithms. $E(S_R)$ has the same units as $S_L$ and $S_U$, thus after calibration (see \S\ref{sec:calibration}) could be given in Jy. In practice, the increase of data noise after correlation is much less severe because of RFI flagging, which excises a part of the RFI. One can assume that all RFI above a certain power level is found by the detector. Since modern RFI detection algorithms can find all RFI that is detectable ``by eye'' \citep{post-correlation-rfi-classification}, this power level will be near the level of the noise mode. We use the AOFlagger for RFI detection, which will be described in Sect.~\ref{sec:dist-data}. In
\citet{lofar-radio-environment}
the false-positives rate for the AOFlagger is estimated to be 0.5\%.

An estimate of $S_d$, the average lower limit of detected RFI, can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the real number (true positives) of RFI samples. Therefore, the limit is calculated similar to Eq.~\eqref{eq:lower-limit-1}, where the term $N_\textrm{part} - N_\textrm{total}$ needs to be replaced with $N_\textrm{RFI}$, which equals the total number of samples detected as RFI minus the false positives.

@@

To detect RFI, the AOFlagger \citep{LOFAR-RFI-pipeline} is used. This RFI detector estimates the sky contribution by iteratively applying a high-pass filter to the visibility amplitudes of a single baseline in the time-frequency plane. Subsequently, it flags line-shaped features with the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. Finally, the scale-invariant rank operator, a morphological technique to search for contaminated samples, is applied on the two-dimensional flag mask \citep{scale-invariant-rank-operator}.

Because the AOFlagger detector is partly amplitude-based, it is likely that low-level RFI will leak through the detector. Since it is also low-level RFI we are interested in, we will analyse unflagged data and the RFI classified data.
